Exploring TTC Delays

### 2024-01-25 outputs/paper/paper.qmd
finished plot code. organized paper.

---
title: "Exploring TTC Delays"
subtitle: "A Multimodal Analysis"
author: Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you. No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook... sigh. I will try the appendix cocktails someday, though. :)"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
fig_caption: yes
number-sections: true
bibliography: references.bib
---

@@

library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(data.table)
library(dplyr)
library(kableExtra)
library(knitr)
```

@@

This paper aims to find patterns of public transit delay time across various genres of transportation. First, the data source and analysis employed are covered in Section Two. Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration for this data.

@@

To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt]. Then, the data was cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages. After combining the datasets together, variables common between datasets and relevant to the analysis were selected, leading to a dataset with 93,569 observations. In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_data
#| tbl-cap: Sample of Cleaned Delay Data
#| fig-align: center

@@

@@

```{r}
#| message: false
#| echo: false
#| label: summary_stats_table
#| tbl-cap: Sample of Cleaned Delay Data
#| fig-align: center

sample_table <- cleaned_transit_data |>
  rename(Vehicle = vehicle) |>
  rename(`Delay (Minutes)` = `Min Delay`) |>
  rename(Reason = Incident) |>
  rename(`Route/Line` = Line) |>
  arrange(Date)

sample_table <- head(sample_table, 5)

#Scaling taken from: 
#https://stackoverflow.com/questions/56254631/table-way-too-wide-to-fit-in-markdown-generated-pdf
kable(sample_table) |> kable_styling(latex_options = c("striped", "scale_down"))
```

```{r}
#| message: false
#| echo: false
#| label: common-reasons-delay
#| tbl-cap: Most Common Reasons for Delay on TTC

cleaned_transit_data |>
  count(Incident) |>
  ggplot(aes(x = n, y = reorder(Incident, n))) +
  geom_bar(stat="identity", fill="grey50") +
  labs(title="Most Common Reasons for Delay", x="Count", y="Incident Type") +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5),
        panel.background = element_blank(),
        axis.line.y=element_line(color="black"),
        axis.ticks.y=element_blank(),
        panel.grid.major.x=element_line(color="grey80"),
        panel.grid.minor=element_blank())

```{r}
#| message: false
#| echo: false
#| label: monthly-delay-times
#| tbl-cap: Average Monthy Delay Times
#Line chart of how avg. delay time changes over the months

month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)

# #Calculate avg. delays 
avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

# #convert month to factor (categorical data)
avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])

# #plot the graph
avg_delay |>
  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
  geom_line() +
  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay") +
  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
  theme_minimal()
```

```{r}
#| message: false
#| echo: false
#| label: mode-longest-delays
#| tbl-cap: Most Common Reasons for Delay on TTC
#Bar graph of mode of transport with longest delays

@@

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```

```{r}
#| message: false
#| echo: false
#| label: most-incidents-vehicle
#| tbl-cap: Most Incidents by Vehicle
#Bar graph of mode of transport with most number of delays

@@

```{r}
#| message: false
#| echo: false
#| label: transport-delay-time
#| tbl-cap: Most Common Reasons for Delay on TTC
#Bar graph of mode of transport with most number of delays

@@

clean_transit_data <- cleaned_transit_data

@@

# Convert 'Day' column to a factor with specific levels
clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)

# Now create the plot
num_incidents_vehicle <- clean_transit_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")

num_incidents_vehicle |>
  ggplot(mapping = aes(x = Day, y = num_incidents)) +
  geom_col() +
  labs(title="Count of Delays by Day of Week", x="Day of Week", y="Incident Count") +
  theme_minimal()

```{r}
#| message: false
#| echo: false
#| label: transport-delay-number
#| tbl-cap: Most Common Reasons for Delay on TTC
#| 
# Define the order of days
days_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

# Convert 'Day' column to a factor with specific levels
clean_transit_data$Day <- factor(clean_transit_data$Day, levels = days_order)

avg_time_delayed_vehicle <- clean_transit_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = Day, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```

@@

@@

>>>>

---
title: "Exploring TTC Delays"
subtitle: "A Multimodal Analysis"
author: Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/OpenDataToronto. Thank you to Nescafé coffee, Rajan Maghera, Hannah Yu, and Sehar Bajwa for your love and support - could not have done it without you. No thanks go to Rohan in particular for not pointing out all the R he wanted us to learn that was in his textbook... sigh. I will try the appendix cocktails someday, though. :)"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
toc: true
fig_caption: yes
number-sections: true
bibliography: references.bib

format: pdf
---

@@

library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(tidyr)
library(data.table)
library(dplyr)
library(kableExtra)
library(knitr)
library(gridExtra)
```

@@

Patterns of public transit delay time are found across various genres of transportation. First, the data source and analysis employed are covered in Section Two. Section Three then critically examines the data, delving deeper to arrive at various conclusions, as well as discussing implications and proposing new areas of exploration for this data. The difference between bus, streetcar, and subway delays is shown to be stark. This analysis highlights the need to focus on especially reducing bus delays on several key routes.

@@

```{r data import}
#| message: false
#| echo: false
#| label: import data
#| tbl-cap: Sample of Cleaned Delay Data
#| fig-align: center

@@

#make each Line a category, not an integer
cleaned_transit_data <- cleaned_transit_data |> mutate(Line = factor(Line))
```

```{r sample table }
#| message: false
#| echo: false
#| label: sample_table
#| tbl-cap: Sample of Cleaned Delay Data
#| fig-align: center

sample_table <- cleaned_transit_data |>
  rename(Vehicle = vehicle) |>
  rename(`Delay (Minutes)` = `Min Delay`) |>
  rename(Reason = Incident) |>
  rename(`Route/Line` = Line) |>
  arrange(Date) |>
  head(5)

#Scaling taken from: 
#https://stackoverflow.com/questions/56254631/table-way-too-wide-to-fit-in-markdown-generated-pdf

kable(sample_table) |> kable_styling(latex_options = c("striped", "scale_down"))
```

To investigate transit delays in Toronto, data on bus [@busdata], subway [@subwaydata], and streetcar [@streetcardata] delay incidents for the year 2023 was downloaded using the OpenDataToronto R package [@odt] from the Toronto Open Data Catalogue. The data was then cleaned using R [@citeR], tidyverse [@tidyverse] and it's associated packages. After combining the datasets together, variables common between datasets and relevant to the analysis were selected, leading to a dataset with 93,569 observations. In the case of this analysis, the variables included were date (year, month, day), day of the week, Location of Delay Incident, Cause of Delay, the time delay (in minutes), Vehicle (bus, subway, streetcar), as well as the route/line (simplified to 'line') the vehicle was on when the incident happened.

## Incident Type {#subsec1-data}
```{r ggplot Common Delay Reasons}
#| message: false
#| echo: false
#| label: common-reasons-delay
#| tbl-cap: Most Common Reasons for Delay on TTC

delay_dataset <- cleaned_transit_data

#Drop N/A incidents (they only apply to subways; subway dataset didn't indicate reasons with entries)
delay_dataset <- delay_dataset |> filter(Incident != "N/A")

delay_dataset |>
  count(Incident) |>
  ggplot(aes(x = n, y = reorder(Incident, n))) +
  geom_bar(stat="identity", fill="grey50") +
  labs(title="Bus and Streetcar Delay Reasons", x="Count", y="Incident Type") +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5, vjust = 0.5),
        panel.background = element_blank(),
        axis.line.y=element_line(color="black"),
        axis.ticks.y=element_blank(),
        panel.grid.major.x=element_line(color="grey80"),
        panel.grid.minor=element_blank(),
        plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))

```
Subways didn't have any reason for delay, yet have the most incidence of delay.

## Line/Route {#subsec2-data}
```{r ggplot top 10 routes with longest avg. delay}
#| message: false
#| echo: false
#| label: lines-longest-delays
#| tbl-cap: Incidence of Delay, by Route/Line

delay_by_line <- cleaned_transit_data

delay_by_line <- delay_by_line |> 
                 group_by(Line) |> 
                 summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop") |> 
                 arrange(desc(avg_delay)) |> 
                 head(10)

# Plotting
ggplot(delay_by_line, aes(x = reorder(Line, desc(avg_delay)), y = avg_delay)) +
  geom_bar(stat = "identity") +
  labs(x = "Route/Line", y = "Average Delay (Minutes)", title = "Top 10 Routes with Longest Average Delay") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))

```
```{r ggplot top 10 routes with most delay incidence}
#| message: false
#| echo: false
#| label: lines-most-delays
#| tbl-cap: Incidence of Delay, by Route/Line

delay_by_line <- cleaned_transit_data

delay_by_line <- delay_by_line |> 
                 group_by(Line) |> 
                 summarise(num_delays = n(), .groups = "drop") |> 
                 arrange(desc(num_delays)) |> 
                 head(10)
delay_by_line

# Plotting
ggplot(delay_by_line, aes(x = reorder(Line, desc(num_delays)), y = num_delays)) +
  geom_bar(stat = "identity") +
  labs(x = "Route/Line", y = "Delay Incidence Count", title = "Top 10 Routes with Highest Delay Incidence") +
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5))

@@

## Vehicle Type {#subsec3-data}
```{r ggplot transport with longest avg. delays}
#| message: false
#| echo: false
#| label: mode-longest-delays
#| tbl-cap: Most Common Reasons for Delay on TTC
#Bar graph of mode of transport with longest delays

@@

avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = vehicle, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay by Vehicle", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```
```{r ggplot transport with most incidence of delays }
#| message: false
#| echo: false
#| label: most-incidents-vehicle
#| tbl-cap: Most Incidents by Vehicle
#Bar graph of mode of transport with most incidence of delays

@@

## Day of Week {#subsec4-data}
```{r ggplot Day of Week with Delay incidence}
#| message: false
#| echo: false
#| label: transport-delay-time
#| tbl-cap: Most Common Reasons for Delay on TTC
#Day of week with most incidence of delay

@@

dow_data <- cleaned_transit_data

@@

# Convert 'Day' column to a factor with specific levels
dow_data$Day <- factor(dow_data$Day, levels = days_order)

# Now create the plot
num_incidents_vehicle <- dow_data |> arrange(Day) |> group_by(Day) |> summarise(num_incidents = n(), .groups = "drop")

num_incidents_vehicle |>
  ggplot(mapping = aes(x = Day, y = num_incidents)) +
  geom_col() +
  labs(title="Incidence of Delay by Day of Week", x="Day of Week", y="Incident Count") +
  theme_minimal()

```
```{r ggplot Day of Week with Avg. Delays}
#| message: false
#| echo: false
#| label: transport-delay-number
#| tbl-cap: Most Common Reasons for Delay on TTC
#|

dow_data <- cleaned_transit_data

# Convert 'Day' column to a factor with specific levels
dow_data$Day <- factor(dow_data$Day, levels = days_order)

#create tibble with avg delay time
avg_time_delayed_vehicle <- dow_data |> group_by(Day) |> summarise(avg_delay = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

#plot
avg_time_delayed_vehicle |>
  ggplot(mapping = aes(x = Day, y = avg_delay)) +
  geom_col() + 
  labs(title="Average Delay Time by Day of Week", x="Vehicle Type", y="Average Delay (mins)") +
  theme_minimal()
```

## Time of Year {#subsec5-data}
```{r ggplot avg delay change monthly}
#| message: false
#| echo: false
#| warning: false
#| label: monthly-delay-times
#| tbl-cap: Average Monthy Delay Times
#Line chart of how avg. delay time changes over the months

# create tibble with month col
month_dataset <- cleaned_transit_data |> mutate(month(Date)) |> rename(cleaned_transit_data$month, `month` = `month(Date)`)

#Calculate avg. delays 
avg_delay <- month_dataset |> group_by(month, vehicle) |> summarise(`Min Delay` = mean(`Min Delay`, na.rm = TRUE), .groups = "drop")

# convert month to factor (categorical data)
avg_delay$month <- factor(avg_delay$month, levels = 1:12, labels = month.abb[1:12])

# plot the graph
avg_delay |>
  ggplot(aes(x=month, y=`Min Delay`, color=vehicle, group=vehicle)) +
  geom_line() +
  labs(title="Average Monthly Delays by Vehicle", x="Month", y="Monthly Avg. Delay (minutes)") +
  scale_color_manual(values=c("blue","green", "red"), labels=c("Bus","Streetcar", "Subway")) +
  theme_minimal()
```

@@

## Results {#subsec1-discussion}

## Implications {#subsec2-discussion}

## New Areas of Exploration {#subsec3-discussion}

# Conclusion

@@

# Appendix: TTC Route Nomenclature
